\vspace{5cm} \normalsize \textsc{\NetIDb} \\ [2.0cm] \HRule{0.5pt} \\ \LARGE \textbf{\uppercase{\documenttitle}} \HRule{2pt} \\ [0.5cm] \normalsize \textsc{Sergio Castañeiras Morales\\NIU:1598456\\ }\normalsize \vspace*{5\baselineskip}

### 2025-05-28 Manuscript/Plantilla memòria TFG 2024_25.tex
miquel's moreno corrections

\newacronym{cox2}{COX-2}{Cyclooxygenase-2}
\newacronym{nsaid}{NSAID}{Non-Steroidal Anti-Inflammatory Drug}
\newacronym{ptgs2}{PTGS2}{Prostaglandin-endoperoxide synthase 2}
\newacronym{ai}{AI}{Artificial intelligence}
\newacronym{ml}{ML}{Machine Learning}
\newacronym{rf}{RF}{Random Forest}
\newacronym{id}{ID}{Identificator}
\newacronym{ic50}{IC$_{50}$}{Half Maximal Inhibitory Concentration}
\newacronym{ic90}{IC$_{90}$}{90 Percent Inhibitory Concentration}
\newacronym{ic99}{IC$_{99}$}{99 Percent Inhibitory Concentration}
\newacronym{tp}{TP}{True Posive}
\newacronym{fp}{FP}{False Posive}
\newacronym{tn}{TN}{True Negative}
\newacronym{fn}{FN}{False Negative}

@@

The main objective of the project is to develop software capable of generating AIs that can predict the 50\% inhibition concentration (IC$_{50}$) for COX-2\footnote{In reality, the software works for any protein with an entry in the ChEMBL database \cite{ChemblDatabase}, although the study focuses on COX-2.} with the highest possible accuracy. To achieve this, all known molecular data with an established inhibition potential for COX-2 are extracted. After a user-configurable filtering process, 5,900 chemical descriptors are calculated for each entry using the AlvaDesk software \cite{MauriMolecularDescriptorsBook}\cite{AlvaDescSecondPaper}. Subsequently, part of the data is used to train \emph{Random Forest} (RF) models \cite{MachineLearningRandomForest}, while the rest is reserved to validate the accuracy of the predictions.

@@

Although significant advancements have been made, cancer still accounts for over 8 million deaths per year worldwide, and the scientific and medical communities remain far from achieving its complete eradication. Inflammation is one of the hallmarks of carcinogenesis, in fact, various cancer therapies target inflammation as a means of preventing and reducing cancer occurrences. When a tissue is damaged, inflammation protects the organism from infections caused by external pathogens, a key function of the immune system to prevent the presence of invaders in the body. During the inflammatory process, cells proliferate under the command of the immune system to replace the damaged cells of the affected tissue. However, if this cell reproduction continues beyond the healing of the damaged tissue, it can potentially lead to cancer, contradicting the initial healing purpose of the inflammatory process. In some cases, inflammation can become chronic, leading to tumor development and uncontrolled cell proliferation. As a result, a wide range of drug prototypes have been designed to suppress inflammation. However, many of these drugs have been linked to severe side effects, including immunosuppression, cardiovascular risks, and gastrointestinal complications. Consequently, the administration of these drugs is often contrindicative, and the search for a more effective and safer treatment remains ongoing. Plently of the research in this field is mainly focused on the pro-inflamatory enzyme \gls{cox2}, one of the main commanders in the inflammation process and responsable to convert the arachidonic acid to prostaglandin \ch{H_2} (Scheme (\ref{Cox2MainReaction})). Subsequently, a therapy based on the chemical inhibition of the \gls{cox2} with no side effects has been one of the research lines in cancer treatment leading to a considerable amount of experiments and data.
\begin{figure}[t]
	\centering
	\includegraphics[width = \textwidth, trim={10cm 0cm 10cm 0cm}, clip]{GeneralSources/Cox2ChimeraX.jpg}
	\caption{Structural representation of the \gls{cox2} generated by the ChimeraX software\cite{ChimeraX}.}
	\label{FigureCox2ChimeraX}
\end{figure}

@@

Determining an experimental \gls{ic50} value can be both costly and time-consuming\footnote{Usually this parameter is computed throughout the Cheng Prusoff Equation (Def. (\ref{definitionChengPrusoff})) with experimental data.}. On the other hand, \gls{ai} provides an alternative by offering highly accurate predictions based on existing data, optimising research processes, and accelerating scientific discovery. In this scenario this project aims to implement artificial intelligence in computational chemistry, concretely, using \gls{ai}-based algorithms to predict a drug’s inhibition potential\cite{BookIC50} for a given protein with a relativelly good accuracy \footnote{The accuracy of the ML models is discussed on the Results and Discussion section (\ref{sectionResultsAndDiscussion}).}. To achieve this, we make use of the ChEMBL database\cite{ChemblDatabase}, a vast repository of bioactive molecules with drug-like properties. We extract all known molecular data with a documented inhibition potential for the target protein, creating a comprehensive dataset. The chemical descriptors (Def. (\ref{definitionChemicalDescriptors})) of each molecule in the database are then computed using AlvaDesk\cite{MauriMolecularDescriptorsBook}\cite{AlvaDescSecondPaper} software. Around 5000 descriptors are calculated\cite{DescriptorsBook}, which comprehend from the elemental molecular weight to the complex equipotential electronic surface, providing critical information about each compound’s behaviour. The resulting dataset is subsequently used to train \gls{ai} models, enabling them to predict the inhibition potential of unknown compounds. Finally, we evaluate the reliability of each model by testing it against real experimental data.

@@

Currently, the pharmacological therapies for the \gls{cox2} inhibition are based on \gls{nsaid}\cite{nsaidDefinition} such as the popular ibuprofen or aspirin. However they have proved to be related to cardiovascular diseases and are contraindicated for people with more than 50 years or people with gastrointestinal problems, among others. Still some alternative therapies related with \gls{nsaid}s are being explored, in particular, one of the most revolutionary ideas is the application of azophotoswitches in drug design.

We define a molecule as an azophotoswitch if it contains an azo bond (\ch{N=N}) that is sensitive to configurational transformations upon photo-excitation. For instance, let us consider the case of (E/Z)-N,1-diphenylmethanimine,
\begin{figure}[H]
\captionsetup{type=scheme}
\centering
\schemestart
\scalebox{0.8}{
	\chemfig[angle increment=30]{*6(-=-(-[1]N=[-1]N-*6(-=-=-=))=-=)}}
	\arrow{<=>[$h\nu$]}
\scalebox{0.8}{
	\chemfig[angle increment=30]{
	N(-[-4]*6(-=-=-=))=[0]N-[-2]*6(-=-=-=)
	}}
\schemestop
	\caption{(E/Z)-N,1-diphenylmethanimine conversion as an example of an azophotoswitch.}
\label{AzophotoswitchExampleFigure}
\end{figure}
The photo-induced E/Z isomerization leads to distinct bioactivities for each configuration. This variation arises from the stereochemical constraints required for a molecule to bind to a target protein. Typically, a protein’s binding site has a specific shape, and only molecules whose conformation matches this shape can interact effectively. In the context of an azophotoswitch, one isomer may fit precisely into the binding pocket, leading to strong interactions, while the other may not. Accordingly, we refer to the isomer that interacts most effectively with the protein as the active configuration, and the other as the inactive configuration.\footnote{In the context of this project, interaction refers to inhibition of the protein’s activity.} \footnote{Generally, the trans-isomer is considered the active configuration, while the cis-isomer is considered inactive. However, exceptions exist.}

@@

The project specifically focuses on the \gls{cox2} protein. Thanks to the availability of a large \gls{ic50} dataset for this target, the model’s predictive reliability can be effectively validated. The prediction process is powered by an \gls{ai} tool designed to deliver high-quality results, taking into account relevant molecular descriptors. The software also serves to cross-check computational results previously obtained by the Physical Chemistry Unit of the UAB \cite{UAB_ComputationalChemistry}. The application centers on azophotoswitches—molecules of particular pharmacological interest discussed in Section \ref{sectionIntroduction}. Additionally, a molecular docking study is performed on one of the most promising azophotoswitch candidates.

@@

\newpage
\section{Methodology}\label{sectionMethodology}
The source code is all stored in the \href{https://github.com/SirSergi0/Repository---AI-application-for-azophotoswitches-optimization-with-pharmacological-interest}{\emph{AI application for azophotoswitches optimization with pharmacological interest} GitHub repository}\cite{GitHub}. The consultace of this repository is highly recommended in order to comprehended the insights in the workflow of the project. By the time this project is presented, it remains as a open-source repository for the Computational Chemistry UAB unit \cite{UAB_ComputationalChemistry}.

The target protein's ID is set at \emph{CHEMBL230} corresponding to the COX-2 ID in the ChEMBL database\cite{ChemblDatabase}. Utilising \emph{requests} python package\cite{PythonPackageRequests} a query URL is sent asking for all molecules with a know $IC_{50}$ value (Def. \ref{definitionIC50}) with a limit of 1000 entries per request. The process is iterated until all data is extracted leading a total of 7979 molecules. Hence the datasheet is processed in pandas dataframes\cite{PythonPackagePandas} and encrypted into binary feather files to optimise reading-writting speed. By removing entries with the same canonical smiles a total of 5112 molecules remain. Among this entries well known drugs such as Celecoxib (Def. \ref{definitionCelecoxib}), Rofecoxib (Def. \ref{definitionRofecoxib}) or even Ibuprofen can be found. However the $IC_{50}$ molecules range is comprehended from $10^{-3}$ to $10^{8}$ nM, a counterproductive range for the AI training procedure. Since we are interested in testing azophotoswitches with presumably low \gls{ic50}, training \gls{ml} models with data of the order of $10^{8}$ or $10^{3}$ nM can be counterproductive since the model might misinterpret the data.\footnote{Typically the range of the azophotoswitches \gls{ic50} is arround the Celecoxib's \gls{ic50}, i.e. arround $120$ nM}. Consecutivelly, a hard-coded range is filtered discarding all molecules outside the given range, for the most part of the analysis this range is set at $[0,300]$ nM \footnote{this $IC_{50}$ working range is the standard in this kind of studies \cite{MachineLearningPaper5Lipoxygenase}.} which reduces the dataset to 1438 entries, i.e. molecules.

With the AlvaDesk-python \cite{AlvaDescSecondPaper} facility, the chemical descriptors (i.e., the chemical fingerprint) of each molecule are computed, providing a total of $5800$ descriptors per molecule. Still, some chemical descriptor need from the presence of a certain atom, for instance a chemical descriptor related to presence of a metal in the molecule, or active gorup, for example chemical descriptor related to the presence of a carboxilic acid, or chemical descriptor related triple bound among other possiblities. The computation of this chemical descriptors appear as \emph{null values}. By deleting molecular descriptors with null values 2917 chemical descriptors remain. Here, the Pearson correlation coefficient (Def. \ref{definitionPersonCoefficient}) between each chemical descriptor and the $IC_{50}$ value is computed, providing insight into the direct relationship between $IC_{50}$ and the descriptors. This relationship will be discussed in the Results and Discussion Section (\ref{sectionResultsAndDiscussion}).

At this stage, the average $IC_{50}$ is calculated, and the neighborhood size corresponding to the percentage defined by the hard-coded variable \emph{percentageErased} is removed. This allows us to distinguish between \emph{highly active molecules} and \emph{least active molecules}, those with lower and higher $IC_{50}$ values, respectively. By doing so, it is possible to compute the classification accuracy statistics of the model from the cluster association of each prediction. Thus we can compare each prediction of each molecules to experimental data, i.e. the model predicts a molecule to be highly active molecule to a highly active molecule this will be denoted as a \gls{tp}. Similarly we can define a \gls{tn}, \gls{fp} and \gls{fn} . This procedure is an standard for evaluating a \gls{ml} model realiability.

@@

As described in definition (\ref{definitionDecisionTree}), the Random Forest algorithm generates a collection of random decision trees. The number of trees in the forest is a configurable parameter of the algorithm. More specifically, the Random Forest algorithm produces a set of decision trees, each trained on a random subset of the training data. The structure of each tree adjusts during training to make predictions that closely approximate the true target values. The final output of the Random Forest is typically an aggregate of the predictions from all the individual trees, such as a majority vote for classification tasks or an average for regression tasks. \par

@@

The \gls{rf} algorithm generates a large number of random decision trees (as defined by the hard-coded variable \emph{numberOfTrees}), which are trained on subsets of the training data. Typically, increasing the number of trees enhances the model’s predictive accuracy, however, this comes at the cost of greater computational complexity. Importantly, prediction accuracy cannot be arbitrarily improved simply by increasing the number of trees. The model’s performance is fundamentally limited by the quality and quantity of the training data. If the data lacks from information about the underlying protein-related phenomena, higher accuracy cannot be achieved regardless of how many trees are used.

In general, the relationship between data quality/quantity and the number of trees can be summarized as the following
\begin{align*}
\uparrow\text{Data} + \uparrow\text{Trees} &\Rightarrow \uparrow\text{Prediction Accuracy} \\
\downarrow\text{Data} + \downarrow\text{Trees} &\Rightarrow \downarrow\text{Prediction Accuracy}
\end{align*}
However, increasing either component in isolation, i.e. add more data or more trees, does not necessarily lead to better accuracy. For the available dataset in this work, we observed training with 500 trees yields similar results to training with 300 trees.

Subsequently, these models are evaluated by predicting the $IC_{50}$ values of the \emph{testing sets}. Using the predictions, we compute the \emph{True Positive Rate} (Def. \ref{definitionTruePositiveRate}), \emph{True Negative Rate} (Def. \ref{definitionTrueNegativeRate}), \emph{Classification Accuracy} (Def. \ref{definitionClassificationAccuracy}), and \emph{Matthews Correlation Coefficient} (Def. \ref{definitionMatthewsCorrelationCoeficient}). Based on these metrics, the variables \emph{percentageErased}, \emph{testSizeProportion}, and \emph{numberOfTrees} are manually adjusted to optimize model performance.

Finally, the \gls{rf} model with the best performance\footnote{The criteria for determining the model with “better statistics” is discussed in Section \ref{sectionPersonCorrelationFactor} since this term references \gls{ai} statistics rather than the usual statistics.} is used to predict the \gls{ic50} values of the azophotoswitch prototypes listed in Appendix \ref{appendixAzophotoshiches}. This process begins with computing the chemical descriptors using AlvaDesc software \cite{MauriMolecularDescriptorsBook}. Irrelevant descriptors are then removed, retaining only those used during training. By inputting the selected descriptors into the trained \gls{rf} model, the predicted \gls{ic50} values are obtained and subsequently analyzed.

Finally, a docking study was performed using GOLD2022 \cite{GoldReference}. This software employs a genetic algorithm, along with molecular modeling techniques, to predict the optimal binding pose of the ligand within the \gls{cox2} binding site. The genetic nature of the algorithm allows for a broad exploration of the conformational and positional space between the ligand and the protein. By evaluating and ranking ligand poses based on predicted interactions, such as hydrogen bonding, van der Waals forces, and metal coordination, unstable or energetically unfavorable conformations are discarded, resulting in a set of plausible, energetically stable binding orientations. The docking calculations were carried out within a 26 Å radius sphere centered at Arg513.

@@

Once the models have been generated, i.e., once the relevant hyperparameters are set, we proceed with generating predictions. At this stage, we compute the Pearson Correlation Coefficient (Def. (\ref{definitionPersonCoefficient})) and the Mean Squared Error (Def. (\ref{definitionMeanSquaredError})) between the predicted and experimental \gls{ic50} values.

@@

More reliable indicators of model's performance are provided by the \emph{True Positive Rate} (Def. \ref{definitionTruePositiveRate}), \emph{True Negative Rate} (Def. \ref{definitionTrueNegativeRate}), \emph{Classification Accuracy} (Def. \ref{definitionClassificationAccuracy}), and \emph{Matthews Correlation Coefficient} (Def. \ref{definitionMatthewsCorrelationCoeficient}), the standard metrics used to evaluate the performance of a \gls{ml} model. Generally speaking, we define a method A to have "better statisitcs" than other method B if the computed quantities: \emph{True Positive Rate}, \emph{True Negative Rate}, \emph{Classification Accuracy}, and \emph{Matthews Correlation Coefficient}, for A take higher values than the same quantities for B.\footnote{Since all quantities are highly bounded by 1 we might also affirm, the closer to 1 the statistics values are the more reliable the method is.} Specifically, if this procedure for the method A and B lead to incoherences, i.e. some statistics of A are better than B while other are worst, we take the Matthews Correlation Factor as the most reliable statistical outcome.\footnote{This is the usual procedure in the analysis of \gls{ml} predictions\cite{MachineLearningPaper5Lipoxygenase}.}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\textwidth, trim={1.75cm 0.8cm 2.5cm 2cm}, clip]{../Plots/GbindingVSIC50.pdf}
	\caption{Predicted \gls{ic50} values for azophotoswitches from Appendix \ref{appendixAzophotoshiches}. The x-axis represents $\Delta G_{binding}$, and the y-axis represents the predicted \gls{ic50}.}
	\label{figureGraphicGbindingIC50}
\end{figure}

Furthermore, the performance metrics for this prediction are summarized in Table (\ref{TableStatisitcsMainResults}). All key statistical indicators exceed the 0.90 threshold. Notably, the model achieves a \emph{True Negative Rate} of 0.99 (Def. (\ref{definitionTrueNegativeRate})), indicating that it correctly identifies non-active inhibitors 99\% of the time. The \emph{Classification Accuracy} (Def. (\ref{definitionClassificationAccuracy})) and \emph{Matthews Correlation Coefficient} both approach 1, reinforcing the reliability of this particular \gls{rf} model in distinguishing between active and inactive azophotoswitches.

>>>>

\newacronym{cox2}{COX-2}{Cyclooxygenase-2}
\newacronym{nsaid}{NSAID}{Non-Steroidal Anti-Inflammatory Drug}
\newacronym{ptgs2}{PTGS2}{Prostaglandin-endoperoxide synthase 2}
\newacronym{ai}{AI}{Artificial intelligence}
\newacronym{ml}{ML}{Machine Learning}
\newacronym{rf}{RF}{Random Forest}
\newacronym{id}{ID}{Identificator}
\newacronym{ic50}{IC$_{50}$}{Half Maximal Inhibitory Concentration}
\newacronym{ic90}{IC$_{90}$}{90 Percent Inhibitory Concentration}
\newacronym{ic99}{IC$_{99}$}{99 Percent Inhibitory Concentration}
\newacronym{tp}{TP}{True Posive}
\newacronym{fp}{FP}{False Posive}
\newacronym{tn}{TN}{True Negative}
\newacronym{fn}{FN}{False Negative}
\newglossaryentry{smiles}{name={SMILES},description={\mbox{Simplified Molecular Input Line Entry System}}}
%!TEX encoding = UTF-8 Unicode\newacronym{smiles}{SMILES}{Simplified Molecular Input Line Entry System}

@@

The main objective of the project is to develop software capable of generating AIs that can predict the 50\% inhibition concentration (IC$_{50}$) for COX-2\footnote{Actually, the software works for any protein with an entry in the ChEMBL database \cite{ChemblDatabase}, although the study focuses on COX-2.} with the highest possible accuracy. To achieve this, all known molecular data with an established inhibition potential for COX-2 are extracted. After a user-configurable filtering process, 5,900 chemical descriptors are calculated for each entry using the AlvaDesk software \cite{MauriMolecularDescriptorsBook}\cite{AlvaDescSecondPaper}. Subsequently, part of the data is used to train \emph{Random Forest} (RF) models \cite{MachineLearningRandomForest}, while the rest is reserved to validate the accuracy of the predictions.

@@

Although significant advancements have been made, cancer still accounts for over 8 million deaths per year worldwide, and the scientific and medical communities remain far from achieving its complete eradication. Inflammation is one of the hallmarks of carcinogenesis, in fact, various cancer therapies target inflammation as a means of preventing and reducing cancer occurrences. When a tissue is damaged, inflammation protects the organism from infections caused by external pathogens, a key function of the immune system to prevent the presence of invaders in the body. During the inflammatory process, cells proliferate under the command of the immune system to replace the damaged cells of the affected tissue. However, if this cell reproduction continues beyond the healing of the damaged tissue, it can potentially lead to cancer, contradicting the initial healing purpose of the inflammatory process. In some cases, inflammation can become chronic, leading to tumour development and uncontrolled cell proliferation. As a result, a wide range of drug prototypes have been designed to suppress inflammation. However, many of these drugs have been linked to severe side effects, including immunosuppression, cardiovascular risks, and gastrointestinal complications. Consequently, the administration of these drugs is often contraindicative, and the search for a more effective and safer treatment remains ongoing. Plenty of the research in this field is mainly focused on the pro-inflamatory enzyme \gls{cox2}, one of the main commanders in the inflammation process and responsible to convert the arachidonic acid to prostaglandin \ch{H_2} (Scheme (\ref{Cox2MainReaction})). Subsequently, a therapy based on the chemical inhibition of the \gls{cox2} with no side effects has been one of the research lines in cancer treatment leading to a considerable amount of experiments and data.
\begin{figure}[t]
	\centering
	\includegraphics[width = \textwidth, trim={10cm 0cm 10cm 0cm}, clip]{GeneralSources/Cox2ChimeraX.jpg}
	\caption{Structural representation of the \gls{cox2} generated by the ChimeraX software\cite{ChimeraX}.}
	\label{FigureCox2ChimeraX}
\end{figure}

@@

Determining an experimental \gls{ic50} value can be both costly and time-consuming\footnote{Usually this parameter is computed throughout the Cheng Prusoff Equation (Definition (\ref{definitionChengPrusoff})) with experimental data.}. On the other hand, \gls{ai} provides an alternative by offering highly accurate predictions based on existing data, optimising research processes, and accelerating scientific discovery. In this scenario this project aims to implement artificial intelligence in computational chemistry, concretely, using \gls{ai}-based algorithms to predict a drug’s inhibition potential\cite{BookIC50} for a given protein with a relativelly good accuracy \footnote{The accuracy of the ML models is discussed on the Results and Discussion section (\ref{sectionResultsAndDiscussion}).}. To achieve this, we make use of the ChEMBL database\cite{ChemblDatabase}, a vast repository of bioactive molecules with drug-like properties. We extract all known molecular data with a documented inhibition potential for the target protein, creating a comprehensive dataset. The chemical descriptors (Definition (\ref{definitionChemicalDescriptors})) \footnote{All the definitions are stored in the Appendix (\ref{appendixRellevantDefinitions}) \emph{Relevant definitions}. We recommend its consultance for further information.}  of each molecule in the database are then computed using AlvaDesk\cite{MauriMolecularDescriptorsBook}\cite{AlvaDescSecondPaper} software. Around 5000 descriptors are calculated\cite{DescriptorsBook}, which comprehend from the elemental molecular weight to the complex equipotential electronic surface, providing critical information about each compound’s behaviour. The resulting dataset is subsequently used to train \gls{ai} models, enabling them to predict the inhibition potential of unknown compounds. Finally, we evaluate the reliability of each model by testing it against real experimental data.

@@

Currently, the pharmacological therapies for the \gls{cox2} inhibition are based on \gls{nsaid}\cite{DefinitionCOX2CancerDictionary} such as the popular ibuprofen or aspirin. However they have proved to be related to cardiovascular diseases and are contraindicated for people with more than 50 years or people with gastrointestinal problems, among others. Still some alternative therapies related with \gls{nsaid}s are being explored, in particular, one of the most revolutionary ideas is the application of azophotoswitches in drug design.

We define a molecule as a \emph{photoswitch} if it contains a bond that undergoes a configurational transformation upon photoexcitation. In particular, we take \emph{azophotoswitches} as a class of photoswitches in which the photo-sensitive bond responsible for isomerization is an azo group (i.e. \ch{N=N}). Azophotoswitches represent the most common and widely studied type of photoswitches due to their robust and tunable photochemical properties. For instance, consider the case of \emph{(cis/trans)-N,1-diphenylmethanimine},
\begin{figure}[H]
\captionsetup{type=scheme}
\centering
\schemestart
\scalebox{0.8}{
	\chemfig[angle increment=30]{*6(-=-(-[1]N=[-1]N-*6(-=-=-=))=-=)}}
	\arrow{<=>[$h\nu$]}
\scalebox{0.8}{
	\chemfig[angle increment=30]{
	N(-[-4]*6(-=-=-=))=[0]N-[-2]*6(-=-=-=)
	}}
\schemestop
	\caption{(cis/trans)-N,1-diphenylmethanimine conversion as an example of an azophotoswitch.}
\label{AzophotoswitchExampleFigure}
\end{figure}
The photo-induced cis/trans isomerization leads to distinct bioactivities for each configuration. This variation arises from the stereochemical constraints required for a molecule to bind to a target protein. Typically, a protein’s binding site has a specific shape, and only molecules whose conformation matches this shape can interact effectively. In the context of an azophotoswitch, one isomer may fit precisely into the binding pocket, leading to strong interactions, while the other may not. Accordingly, we refer to the isomer that interacts most effectively with the protein as the active configuration, and the other as the inactive configuration.\footnote{In the context of this project, interaction refers to inhibition of the protein’s activity.} \footnote{Generally, the trans-isomer is considered the active configuration, while the cis-isomer is considered inactive. However, exceptions exist.}

@@

The project specifically focuses on the \gls{cox2} protein. Thanks to the availability of a large \gls{ic50} dataset for this target, the model’s predictive reliability can be effectively validated. The prediction process is powered by an \gls{ai} tool designed to deliver high-quality results, taking into account relevant molecular descriptors. The software also serves to cross-check computational results previously obtained by the Physical Chemistry Unit of the UAB \cite{UAB_ComputationalChemistry}. The application centers on azophotoswitches, molecules of particular pharmacological interest as discussed in Section \ref{sectionIntroduction}. Additionally, a molecular docking study is performed on one of the most promising azophotoswitch candidates.

@@

\newpage
\section{Methodology}\label{sectionMethodology}
The source code is entirely stored in the \href{https://github.com/SirSergi0/Repository---AI-application-for-azophotoswitches-optimization-with-pharmacological-interest}{\emph{AI application for azophotoswitches optimization with pharmacological interest} GitHub repository}\cite{GitHub} and it has been entirely written by Sergio Castañeiras Morales for this project. The consultace of this repository is highly recommended in order to comprehended the insights and the workflow of the project. By the time this project is presented, it remains as an open-source repository of the MolBioMed research group \cite{molbiomed}.

The target protein's ID is set at \emph{CHEMBL230} corresponding to the COX-2 ID in the ChEMBL database\cite{ChemblDatabase}. Utilising \emph{requests} python package\cite{PythonPackageRequests} a query URL is sent asking for all molecules with a know $IC_{50}$ value (Definition \ref{definitionIC50}) with a limit of 1000 entries per request. The process is iterated until all data is extracted leading a total of 7979 molecules. Hence the datasheet is processed in pandas dataframes\cite{PythonPackagePandas} and encrypted into binary feather files to optimise reading-writting speed. By removing entries with the same canonical \gls{smiles} (Definition \ref{definitionSMILES})  a total of 5112 molecules remain. Among this entries well known drugs such as Celecoxib (Definition \ref{definitionCelecoxib}), Rofecoxib (Definition \ref{definitionRofecoxib}) or even Ibuprofen can be found. However the $IC_{50}$ molecules range is comprehended from $10^{-3}$ to $10^{8}$ nM, a counterproductive range for the AI training procedure. Since we are interested in testing azophotoswitches with presumably low \gls{ic50}, training \gls{ml} models with data of the order of $10^{8}$ or $10^{3}$ nM can be counterproductive since the model might misinterpret the data.\footnote{Typically the range of the azophotoswitches \gls{ic50} is arround the Celecoxib's \gls{ic50}, i.e. arround $120$ nM}. Consecutivelly, a hard-coded range is filtered discarding all molecules outside the given range, for the most part of the analysis this range is set at $[0,300]$ nM \footnote{this $IC_{50}$ working range is the standard in this kind of studies \cite{MachineLearningPaper5Lipoxygenase}.} which reduces the dataset to 1438 entries, i.e. molecules.

With the AlvaDesk-python \cite{AlvaDescSecondPaper} facility, the chemical descriptors (i.e., the chemical fingerprint) of each molecule are computed, providing a total of $5800$ descriptors per molecule. Still, some chemical descriptor need from the presence of a certain atom, for instance a chemical descriptor related to the presence of a metal in the molecule, or active gorup, for example chemical descriptor related to the presence of a carboxilic acid, or chemical descriptor related triple bound among other possiblities. The computation of this chemical descriptors appear as \emph{null values}. By deleting molecular descriptors with null values 2917 chemical descriptors remain. Here, the Pearson correlation coefficient (Definition \ref{definitionPersonCoefficient}) between each chemical descriptor and the $IC_{50}$ value is computed, providing insight into the direct relationship between $IC_{50}$ and the descriptors. This relationship will be discussed in the Results and Discussion Section (\ref{sectionResultsAndDiscussion}).

At this stage, the average $IC_{50}$ is calculated, and the neighborhood size corresponding to the percentage defined by the hard-coded variable \emph{percentageErased} is removed. This allows us to distinguish between \emph{highly active molecules} and \emph{least active molecules}, those with lower and higher $IC_{50}$ values, respectively. By doing so, it is possible to compute the classification accuracy statistics of the model from the cluster association of each prediction. Thus we can compare each prediction of each molecules with experimental data. For instance, for an experimentally proved highly active molecule if the model predicts this molecule to be highly active (has a low \gls{ic50} precition) we consider this as \gls{tp}. Similarly we define a \gls{tn}, \gls{fp} and \gls{fn} . This procedure is an standard for evaluating a \gls{ml} model realiability.

@@

As described in Definition (\ref{definitionDecisionTree}), the Random Forest algorithm generates a collection of random decision trees. The number of trees in the forest is a configurable parameter of the algorithm. More specifically, the Random Forest algorithm produces a set of decision trees, each trained on a random subset of the training data. The structure of each tree adjusts during training to make predictions that closely approximate the true target values. The final output of the Random Forest is typically an aggregate of the predictions from all the individual trees, such as a majority vote for classification tasks or an average for regression tasks. \par

@@

The \gls{rf} algorithm generates a large number of random decision trees (as defined by the hard-coded variable \emph{numberOfTrees}), which are trained on subsets of the training data. Typically, increasing the number of trees enhances the model’s predictive accuracy. However, this comes at the cost of greater computational complexity. Importantly, prediction accuracy cannot be arbitrarily improved simply by increasing the number of trees. The model’s performance is fundamentally limited by the quality and quantity of the training data. If the data lacks from information about the underlying protein-related phenomena, higher accuracy cannot be achieved regardless of how many trees are used.

In general, the relationship between data quality/quantity and the number of trees can be summarized as the following
\begin{align*}
\uparrow\text{Data} + \uparrow\text{Trees} &\Rightarrow \uparrow\text{Prediction Accuracy} \\
\downarrow\text{Data} + \downarrow\text{Trees} &\Rightarrow \downarrow\text{Prediction Accuracy}
\end{align*}
However, increasing either component in isolation, i.e. adding more data or more trees, does not necessarily lead to better accuracy. For the available dataset in this work, we observed training with 500 trees yields similar results respect the ones obtained by training with 300 trees.

Subsequently, these models are evaluated by predicting the $IC_{50}$ values of the \emph{testing sets}. Using the predictions, we compute the \emph{True Positive Rate} (Definition \ref{definitionTruePositiveRate}), \emph{True Negative Rate} (Definition \ref{definitionTrueNegativeRate}), \emph{Classification Accuracy} (Definition \ref{definitionClassificationAccuracy}), and \emph{Matthews Correlation Coefficient} (Definition \ref{definitionMatthewsCorrelationCoeficient}). Based on these metrics, the variables \emph{percentageErased}, \emph{testSizeProportion}, and \emph{numberOfTrees} are manually adjusted to optimize model performance.

Finally, the \gls{rf} model with the best performance\footnote{The criteria for determining the model with “better statistics” is discussed in Section \ref{sectionPersonCorrelationFactor} since this term references \gls{ai} statistics rather than the usual statistics.} is used to predict the \gls{ic50} values of the azophotoswitch prototypes listed in Appendix \ref{appendixAzophotoshiches}. This process begins with computing the chemical descriptors using AlvaDesc software \cite{MauriMolecularDescriptorsBook}. Irrelevant descriptors are then removed, retaining only those used during training. By introducing the selected descriptors into the trained \gls{rf} model, the predicted \gls{ic50} values are obtained and subsequently analyzed.

Finally, a docking study was performed using GOLD2022 \cite{GoldReference}. This software employs a genetic algorithm, along with molecular modeling techniques, to predict the optimal binding pose of the ligand within the \gls{cox2} binding site. The genetic nature of the algorithm allows for a broad exploration of the conformational and positional space between the ligand and the protein. By evaluating and ranking ligand interactions, such as hydrogen bonding, van der Waals forces, and metal coordination, unstable or energetically unfavorable conformations are discarded, resulting in a set of plausible, energetically stable binding orientations. The docking calculations were carried out within a 26 Å radius sphere centered at Arg513.

@@

Once the models have been generated, i.e., once the relevant hyperparameters are set, we proceed with generating predictions. At this stage, we compute the Pearson Correlation Coefficient (Definition (\ref{definitionPersonCoefficient})) and the Mean Squared Error (Definition (\ref{definitionMeanSquaredError})) between the predicted and experimental \gls{ic50} values.

@@

More reliable indicators of model's performance are provided by the \emph{True Positive Rate} (Definition \ref{definitionTruePositiveRate}), \emph{True Negative Rate} (Definition \ref{definitionTrueNegativeRate}), \emph{Classification Accuracy} (Definition \ref{definitionClassificationAccuracy}), and \emph{Matthews Correlation Coefficient} (Definition \ref{definitionMatthewsCorrelationCoeficient}), the standard metrics used to evaluate the performance of a \gls{ml} model. Generally speaking, we define a method A to have "better statisitcs" than other method B if the computed quantities: \emph{True Positive Rate}, \emph{True Negative Rate}, \emph{Classification Accuracy}, and \emph{Matthews Correlation Coefficient}, for A take higher values than the same quantities for B.\footnote{Since all quantities are highly bounded by 1 we might also affirm, the closer to 1 the statistics values are the more reliable the method is.} Specifically, if this procedure for the method A and B lead to incoherences, i.e. some statistics of A are better than B while other are worst, we take the Matthews Correlation Factor as the most reliable statistical outcome.\footnote{This is the usual procedure in the analysis of \gls{ml} predictions\cite{MachineLearningPaper5Lipoxygenase}.}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\textwidth, trim={1.5cm 0.6cm 2.4cm 2cm}, clip]{../Plots/GbindingVSIC50.pdf}
	\caption{Predicted \gls{ic50} values for azophotoswitches from Appendix \ref{appendixAzophotoshiches}. The x-axis represents $\Delta G_{binding}$, and the y-axis represents the predicted \gls{ic50}.}
	\label{figureGraphicGbindingIC50}
\end{figure}

Furthermore, the performance metrics of the \gls{rf} model for this predictions are summarized in Table (\ref{TableStatisitcsMainResults}) that yield the results of Tables (\ref{TableMainResults1}) and (\ref{TableMainResults2}) from the appendix (\ref{appendixTableOfResults}) \emph{Tables of results}. All key statistical indicators exceed the 0.90 threshold. Notably, the model achieves a \emph{True Negative Rate} of 0.99 (Definition (\ref{definitionTrueNegativeRate})), indicating that it correctly identifies non-active inhibitors 99\% of the time. The \emph{Classification Accuracy} (Definition (\ref{definitionClassificationAccuracy})) and \emph{Matthews Correlation Coefficient} both approach 1, reinforcing the reliability of this particular \gls{rf} model in distinguishing between active and inactive azophotoswitches.
